---
title: "Tutorial: Scale-informed mixed-effects differential expression with ALDEx3"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Overview

This tutorial demonstrates one scale-informed ALDEx3 mixed-effects analysis
using a small, downsampled subset of the primary MERFISH dataset.

It preserves the manuscript workflow:

- raw counts are stored as genes × cells;
- cells are aligned between counts and metadata;
- genes detected in at least 20% of cells are retained;
- age and cell volume are modeled as fixed effects;
- batch is modeled as a random intercept;
- an informed prior represents an expected 20% reduction in total transcript
  abundance in old cells;
- ALDEx3 propagates compositional and scale uncertainty.

For desktop use, this tutorial uses `method = "lme4"` and 100 Monte Carlo
samples. The manuscript analysis used 2,000 samples to optimize stability and
convergence.

# Expected directory structure

```text
tutorial/
├── tutorial_ALDEx3.Rmd
├── data/
│   ├── tutorial_counts.csv
│   └── tutorial_metadata.csv
└── output/
```

The count matrix must have genes in rows and cells in columns. Its first CSV
column contains gene names, and its first row contains cell identifiers. The
first metadata column must contain the same cell identifiers.

When the R Markdown document is rendered, its relative paths are resolved from
the `tutorial/` directory. The results workbook is therefore written to
`documentation_code/aldex_main_crossstudy_repo/tutorial/output/`. The final
code chunk prints the normalized absolute output path.

# Load packages

Install the tutorial dependencies once from the repository root:

```r
source(
  "documentation_code/aldex_main_crossstudy_repo/tutorial/install_tutorial_dependencies.R"
)
```

The installation script installs the packages used below and pins ALDEx3 to
the exact standard-engine commit used for the manuscript. The historical
reference environment used R 4.5.0 and `lme4` 1.1-37 on Linux. The tutorial
has also completed successfully under R 4.6.1 and `lme4` 2.0.6 on Windows,
but numerical equivalence across those versions has not been established.

```{r packages}
library(ALDEx3)
library(lme4)
library(reformulas)
library(writexl)
library(ggplot2)
```

# Analysis parameters

```{r parameters}
count_file <- file.path("data", "tutorial_counts.csv")
metadata_file <- file.path("data", "tutorial_metadata.csv")
output_dir <- "output"

dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

detection_threshold <- 0.20
S_draws <- 100
gamma_val <- 0.80
scale_sd <- 0.25
ncores <- 1

# Used only to make the tutorial reproducible.
set.seed(42)
```

# Read counts

```{r read-counts}
seq.data <- read.csv(
  count_file,
  row.names = 1,
  check.names = FALSE
)

seq.data <- as.data.frame(seq.data)

dim(seq.data)
seq.data[seq_len(min(5, nrow(seq.data))),
         seq_len(min(5, ncol(seq.data)))]
```

```{r validate-counts}
if (anyNA(seq.data)) {
  stop("The count matrix contains missing values.")
}

if (any(as.matrix(seq.data) < 0)) {
  stop("The count matrix contains negative values.")
}

if (any(as.matrix(seq.data) != floor(as.matrix(seq.data)))) {
  stop("The count matrix contains non-integer values.")
}

if (anyDuplicated(rownames(seq.data))) {
  stop("The count matrix contains duplicated gene names.")
}

if (anyDuplicated(colnames(seq.data))) {
  stop("The count matrix contains duplicated cell identifiers.")
}
```

# Read and prepare metadata

The first metadata column contains cell identifiers. Required columns are
`Age`, `batch`, and `volume`.

```{r read-metadata}
metadata <- read.csv(
  metadata_file,
  check.names = FALSE,
  stringsAsFactors = FALSE,
  colClasses = "character"
)

cell_ids <- metadata[[1]]

if (anyDuplicated(cell_ids)) {
  stop("The metadata contains genuinely duplicated cell identifiers.")
}

rownames(metadata) <- cell_ids
metadata[[1]] <- NULL
metadata[] <- lapply(metadata, type.convert, as.is = TRUE)

required_columns <- c("Age", "batch", "volume")
missing_columns <- setdiff(required_columns, colnames(metadata))

if (length(missing_columns) > 0) {
  stop(
    "Metadata is missing required columns: ",
    paste(missing_columns, collapse = ", ")
  )
}

if (anyDuplicated(rownames(metadata))) {
  stop("The metadata contains duplicated cell identifiers.")
}

metadata$age_binary <- ifelse(metadata$Age == "Yng", 0, 1)
metadata$batch <- factor(metadata$batch)
metadata$volume <- as.numeric(metadata$volume)

head(metadata)
table(metadata$Age, useNA = "ifany")
table(metadata$batch, useNA = "ifany")
```

```{r validate-metadata}
unexpected_age_labels <- setdiff(unique(metadata$Age), c("Yng", "Old"))

if (length(unexpected_age_labels) > 0) {
  stop(
    "Unexpected values in Age: ",
    paste(unexpected_age_labels, collapse = ", ")
  )
}

if (anyNA(metadata$volume)) {
  stop("The volume column contains missing or nonnumeric values.")
}

if (nlevels(metadata$batch) < 2) {
  stop("At least two batch levels are required for the random-intercept model.")
}
```

# Align cells

```{r align-cells}
common_cells <- intersect(colnames(seq.data), rownames(metadata))

if (length(common_cells) == 0) {
  stop("No overlapping cell identifiers were found.")
}

# Preserve count-matrix column order.
common_cells <- colnames(seq.data)[colnames(seq.data) %in% common_cells]

seq.data <- seq.data[, common_cells, drop = FALSE]
metadata <- metadata[common_cells, , drop = FALSE]
metadata <- droplevels(metadata)

stopifnot(identical(colnames(seq.data), rownames(metadata)))

dim(seq.data)
dim(metadata)
```

# Filter genes by detection fraction

```{r filter-genes}
expression_fraction <- rowMeans(seq.data > 0, na.rm = TRUE)
keep_genes <- expression_fraction >= detection_threshold
filt_data <- seq.data[keep_genes, , drop = FALSE]

if (nrow(filt_data) == 0) {
  stop("No genes remain after detection filtering.")
}

data.frame(
  genes_before_filtering = nrow(seq.data),
  genes_after_filtering = nrow(filt_data),
  cells = ncol(filt_data)
)
```

```{r detection-summary}
detection_summary <- data.frame(
  gene = names(expression_fraction),
  detection_fraction = expression_fraction,
  retained = keep_genes,
  row.names = NULL
)

head(
  detection_summary[
    order(detection_summary$detection_fraction, decreasing = TRUE),
  ],
  10
)
```

# Specify the mixed-effects model

The model is:

```text
age_binary + volume + (1 | batch)
```

```{r model-formula}
r_formula <- ~ age_binary + volume + (1 | batch)

fixed_effect_design <- model.matrix(
  reformulas::nobars(r_formula),
  data = metadata
)

if (ncol(fixed_effect_design) > nrow(fixed_effect_design)) {
  stop(
    "The fixed-effect design has more parameters than observations: P = ",
    ncol(fixed_effect_design),
    ", N = ",
    nrow(fixed_effect_design),
    "."
  )
}

colnames(fixed_effect_design)
```

# Construct the informed scale matrix

The informed model assumes that old cells have, on average, 80% of the total
transcript abundance of young cells.

```{r informed-scale}
delta_age <- log2(gamma_val)

theta.perp <- rnorm(
  S_draws,
  mean = delta_age,
  sd = scale_sd
)

logScale_mat <- metadata$age_binary %*% t(theta.perp)

stopifnot(
  nrow(logScale_mat) == ncol(filt_data),
  ncol(logScale_mat) == S_draws
)

dim(logScale_mat)
summary(as.vector(logScale_mat[metadata$age_binary == 0, , drop = FALSE]))
summary(as.vector(logScale_mat[metadata$age_binary == 1, , drop = FALSE]))
```

Young cells receive an offset of zero because `age_binary = 0`. Old cells
receive the sampled age-associated log2-scale offset because
`age_binary = 1`.

# Run ALDEx3

```{r run-aldex, cache=TRUE}
return_pars <- c(
  "X",
  "estimate",
  "std.error",
  "p.val",
  "p.val.adj",
  "logComp",
  "logScale",
  "random.effects"
)

aldex.res <- aldex(
  Y = filt_data,
  X = r_formula,
  data = metadata,
  method = "lme4",
  nsample = S_draws,
  scale = logScale_mat,
  n.cores = ncores,
  return.pars = return_pars
)
```

Because `scale` is supplied directly as a cells × Monte Carlo matrix, `gamma`
is not passed separately.

# Build the results table

```{r build-results}
est_mat <- apply(
  aldex.res$estimate,
  c(1, 2),
  mean,
  na.rm = TRUE
)

se_mat <- apply(
  aldex.res$std.error,
  c(1, 2),
  mean,
  na.rm = TRUE
)

p_mat <- aldex.res$p.val
padj_mat <- aldex.res$p.val.adj

est_t <- t(est_mat)
se_t <- t(se_mat)
p_t <- t(p_mat)
padj_t <- t(padj_mat)

genes <- rownames(est_t)
effects <- colnames(est_t)

results.df <- data.frame(
  gene = genes,
  stringsAsFactors = FALSE
)

for (eff in effects) {
  results.df[[paste0(eff, ":est")]] <- est_t[, eff]
  results.df[[paste0(eff, ":estSE")]] <- se_t[, eff]
  results.df[[paste0(eff, ":pval")]] <- p_t[, eff]
  results.df[[paste0(eff, ":pval.adj")]] <- padj_t[, eff]
}

head(results.df)
```

# Inspect the age effect

```{r age-effect}
age_effect <- effects[grepl("^age_binary", effects)]

if (length(age_effect) != 1) {
  stop(
    "Expected exactly one age coefficient, but found: ",
    paste(age_effect, collapse = ", ")
  )
}

age_estimate_column <- paste0(age_effect, ":est")
age_se_column <- paste0(age_effect, ":estSE")
age_pvalue_column <- paste0(age_effect, ":pval")
age_padj_column <- paste0(age_effect, ":pval.adj")

age_results <- results.df[
  order(results.df[[age_padj_column]], results.df[[age_pvalue_column]]),
  c(
    "gene",
    age_estimate_column,
    age_se_column,
    age_pvalue_column,
    age_padj_column
  )
]

head(age_results, 20)
```

Positive age coefficients indicate higher inferred log abundance in old cells
relative to young cells. Negative coefficients indicate lower inferred
abundance in old cells.

# Plot age effects

```{r plot-age-effects, fig.width=8, fig.height=5}
plot_data <- age_results
plot_data$significant <- plot_data[[age_padj_column]] < 0.05

ggplot(
  plot_data,
  aes(
    x = .data[[age_estimate_column]],
    y = -log10(.data[[age_padj_column]])
  )
) +
  geom_point(aes(shape = significant), alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Old-versus-young ALDEx3 estimate",
    y = expression(-log[10]("adjusted p-value")),
    shape = "Adjusted p < 0.05",
    title = "Scale-informed ALDEx3 age effects"
  ) +
  theme_classic()
```

# Save output

```{r save-output}
output_file <- file.path(
  output_dir,
  "tutorial_aldex3_lme4_results.xlsx"
)

write_xlsx(
  list(
    all_effects = results.df,
    age_effect = age_results,
    gene_filter = detection_summary
  ),
  path = output_file
)

normalizePath(output_file, winslash = "/", mustWork = TRUE)
```

# Session information

```{r session-info}
sessionInfo()
```
